Distillation is widely used to make large language model capabilities cheaper to deploy, and has applications in domain-specific assistants, task-focused reasoners, and other systems where a small specialized model is preferable to serving a much larger general model. The distillation problem is to turn the behavior of a capable teacher into a training signal that a smaller student can actually learn from, despite differences in capacity, output distribution, and inference-time behavior. Recent work has addressed this problem by training students from teacher-generated data, introducing intermediate or adaptive distillation targets, and using on-policy distillation in which a teacher scores rollouts sampled from the current student. Unfortunately, these approaches leave open how distillation behaves when the teacher is hundreds of times larger than the student, a regime where the strongest teachers can also produce supervision that is too complex for the student to learn effectively.

This thesis studies extreme capacity gaps between teachers and students directly and focuses on the on-policy distillation setting. On-policy distillation reduces the mismatch between training and inference-time behavior, but it does not by itself solve the problem that small students may receive teacher signals that are too verbose, too sharp, or too specialized to internalize safely. Consequently, we develop and evaluate on-policy distillation variants for mathematical reasoning, including privileged-context teacher scoring, top-$k$ teacher losses, and student-visible context distillation for encouraging concise reasoning. While promising, this regime is non-trivial because on-policy rollouts must be generated inside the training loop, large teachers must provide token-level feedback efficiently, and specialization can cause catastrophic forgetting outside the target domain. Therefore, we further implement scalable training infrastructure for large-teacher on-policy distillation, release it as a TRL trainer, and evaluate target-domain performance, token efficiency, and out-of-domain retention across capacity gaps of up to 400$\times$. The results show that OPD improves mathematical reasoning, but teacher scaling can reduce gap recovery and harm out-of-distribution reasoning for the smallest student, so useful distillation depends on matching the teacher signal to the student's capacity rather than simply using the largest available teacher. A teacher-assisted privileged-context variant also fails to improve transfer, showing that reducing the apparent capacity gap is not sufficient when the intermediate teacher cannot turn additional context into useful token-level feedback. Overall, the thesis shows that the main bottleneck in extreme-capacity-gap OPD is not access to a stronger or richer teacher signal, but whether that signal is usable for the student on the states it actually visits during training.
