\chapter{Method}

\section{Distillation}

Distillation transfers knowledge from a teacher model to a student model by training the student to match teacher behavior on a distribution of prompts and generated continuations. Following the notation of on-policy generalized knowledge distillation, let $x$ denote a prompt, let $V$ denote the vocabulary, and let $y = (y_1, y_2, \ldots, y_{L_y})$ denote an output sequence. The prefix before token $y_n$ is $y_{<n} = (y_1, \ldots, y_{n-1})$. We write the student policy as $\pi_\theta(\cdot \mid x, y_{<n})$ and the teacher policy as $\pi_T(\cdot \mid x, y_{<n})$, both of which define next-token distributions over $V$.

The simplest distillation objective treats a target sequence as a hard label and trains the student with cross-entropy. If $\mathcal{D}$ is a dataset of prompt--completion pairs, where the completions can be human-written or sampled from the teacher, the loss is

\begin{equation}
\mathcal{L}_{\mathrm{CE}}(\theta)
= \mathrm{E}_{(x,y) \sim \mathcal{D}}
\left[
- \sum_{n=1}^{L_y}
\log \pi_\theta(y_n \mid x, y_{<n})
\right].
\label{eq:xentropy}
\end{equation}

This objective only uses the observed target token $y_n$ at each position. A richer form of distillation instead uses the full teacher distribution over the vocabulary. For a fixed prompt and continuation, the token-level forward KL from teacher to student is

\begin{equation}
D_{\mathrm{KL}}(\pi_T \| \pi_\theta)(y \mid x)
= \sum_{n=1}^{L_y} \sum_{v \in V}
\pi_T(v \mid x, y_{<n})
\log
\frac{\pi_T(v \mid x, y_{<n})}
{\pi_\theta(v \mid x, y_{<n})}.
\label{eq:forward_kl}
\end{equation}

The reverse KL swaps the order of the two distributions:

\begin{equation}
D_{\mathrm{KL}}(\pi_\theta \| \pi_T)(y \mid x)
= \sum_{n=1}^{L_y} \sum_{v \in V}
\pi_\theta(v \mid x, y_{<n})
\log
\frac{\pi_\theta(v \mid x, y_{<n})}
{\pi_T(v \mid x, y_{<n})}.
\label{eq:reverse_kl}
\end{equation}

The distinction is important because KL divergence is asymmetric. Forward KL averages the log-ratio under the teacher distribution, so it penalizes the student for failing to place probability mass on tokens the teacher considers likely. This makes it a mass-covering objective. Reverse KL averages under the student distribution, so it penalizes tokens the student samples when the teacher assigns them low probability. This makes it more mode-seeking and can be useful when a lower-capacity student cannot match the full teacher distribution.

The general on-policy form samples continuations from the student and then compares teacher and student distributions on the prefixes that the student actually visits:

\begin{equation}
\mathcal{L}_{\mathrm{OD}}(\theta)
= \mathrm{E}_{x \sim \mathcal{X}}
\left[
\mathrm{E}_{y \sim \pi_\theta(\cdot \mid x)}
\left[
D(\pi_T, \pi_\theta)(y \mid x)
\right]
\right],
\label{eq:on_policy_distillation}
\end{equation}

where $D$ can be instantiated as the forward KL in Equation~\ref{eq:forward_kl}, the reverse KL in Equation~\ref{eq:reverse_kl}, or another token-level divergence. More generally, generalized knowledge distillation mixes fixed data with on-policy student samples:

\begin{equation}
\begin{array}{rl}
\mathcal{L}_{\mathrm{GKD}}(\theta)
=&
(1-\lambda)\,
\mathrm{E}_{(x,y) \sim \mathcal{D}}
\left[
D(\pi_T, \pi_\theta)(y \mid x)
\right]
\\
&+
\lambda\,
\mathrm{E}_{x \sim \mathcal{X}}
\left[
\mathrm{E}_{y \sim \pi_\theta(\cdot \mid x)}
\left[
D(\pi_T, \pi_\theta)(y \mid x)
\right]
\right],
\end{array}
\label{eq:gkd}
\end{equation}

with $\lambda \in [0,1]$ controlling the fraction of on-policy student-generated data. Setting $\lambda = 0$ recovers distillation on fixed data, while $\lambda = 1$ gives purely on-policy distillation.

In practice, computing the full sums over $V$ can be expensive. The sampled-token reverse-KL formulation used in on-policy distillation keeps only the realized token at each position of a student rollout $\hat{y} \sim \pi_\theta(\cdot \mid x)$. The resulting top-1 sampled-token estimate is

\begin{equation}
\widehat{D}_{\mathrm{KL}}^{\mathrm{top}\mbox{-}1}
(\pi_\theta \| \pi_T)(\hat{y} \mid x)
= \sum_{n=1}^{L_{\hat{y}}}
\left[
\log \pi_\theta(\hat{y}_n \mid x, \hat{y}_{<n})
- \log \pi_T(\hat{y}_n \mid x, \hat{y}_{<n})
\right].
\label{eq:sampled_token_reverse_kl}
\end{equation}

This is the sampled-token analogue of Equation~\ref{eq:reverse_kl}: rather than matching the full next-token distributions, it scores the token actually produced by the student under both policies.

\section{Types of Distillation}

Distillation as a task is generally defined as extracting knowledge from a teacher to a student model, but the actual process can take multiple forms depending on the source of the data we learn from and when we produce the teacher feedback for the student.

Starting from when we produce the feedback from the teacher, a distillation method can be online or offline. In offline methods, you 

\section{Top-K KL}

The equation in 

\section{Privileged Context}
