\chapter{Related Work}

\section{Distillation}

A useful line of prior work on the capacity gap in knowledge distillation focuses on modifying the transfer path or target so that a small student is not forced to directly mimic a much larger teacher. Teacher Assistant Knowledge Distillation \cite{mirzadehImprovedKnowledgeDistillation2019} is one of the clearest early formulations of this problem. It shows that distillation quality can degrade when the teacher is too large relative to the student, and proposes inserting one or more intermediate-sized teacher assistants to bridge that gap through multi-step distillation. Zhang et al. \cite{zhangLiftingCurseCapacity2023} address the same issue by increasing the student’s effective learning capacity with a sparse MiniMoE design. This design adds parameters while keeping inference cost nearly unchanged, so the student can absorb more of the teacher’s knowledge. More recently, TAID: Shing et al \cite{shing_taid_2025} introduce an adaptive intermediate target that gradually shifts from the student’s current distribution toward the teacher’s, making the transfer process smoother over training. Taken together, these works frame the capacity gap as a central bottleneck in distillation and suggest three complementary remedies: bridge the gap with intermediate models, increase the student’s effective capacity, or soften the teacher target through adaptive interpolation.

\section{On-Policy Distillation}

On-policy distillation extends standard distillation by moving supervision from teacher-generated trajectories to student-generated trajectories, and therefore addressing the distribution mismatch that arises when a student is only trained on states visited by the teacher. This shift is formalized in \cite{agarwal_-policy_2024}, which trains on the student’s own outputs while querying the teacher for feedback. More recent work shows how using the teacher’s log-probabilities on student rollouts to define a per-token reverse-KL penalty provides dense feedback at the exact states the student visits \cite{lu_-policy_2025}.

\section{Distillation with Large Capacity Gap}

\section{Using Privileged Information}

\textbf{Teachers with privileged information about the task.} Recent work in self-distillation \cite{shenfeldSelfDistillationEnablesContinual2026, hubotterReinforcementLearningSelfDistillation2026, penalozaPrivilegedInformationDistillation2026, zhaoSelfDistilledReasonerOnPolicy2026} showed the effectiveness of using in-context learning capabilities of teachers to improve the distillation process. The results show that it is possible to turn a model into an effective teacher by providing privileged information during training. This line of work focuses on bootstrapping the student model with a teacher that is initially the same as the student, but they do not explore how their in-context learning techniques perform when the teacher and the student are different models.
